Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-30 paper/main.tex
implemented some of @jeremymanning's edits -- thank you!

We designed Davos to occupy a ``sweet spot'' within this spectrum.
Davos is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
Davos-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run a standard Jupyter
notebook. And like virtual environments, Davos provides a convenient
mechanism for fully specifying (and installing, as needed) a complete set of
Python dependencies, including specific package versions, which are contained
and isolated from the rest of the user's system.

@@

\item \texttt{.all\_projects}: This attribute contains a list of all Davos projects that exist on the user's local system (see Sec.~\ref{subsec:projects} for more information about Davos projects).
  Each item in this list is either a \texttt{Project} or \texttt{AbstractProject} instance.
  \texttt{AbstractProject}s represent notebook-specific projects whose associated notebooks no longer exist.
  They support the same functionality as \texttt{Project} objects (including methods for inspecting, renaming, and deleting them) and serve primarily to help users identify and clean up extraneous projects left behind after deleting Davos-enhanced notebooks (e.g., see Sec.~\ref{subsec:toplevel}).

@@

After re-enabling these two options (line 24), we next \texttt{smuggle}
specific versions of three plotting packages:
\texttt{Matplotlib}~\cite{Hunt07}, \texttt{seaborn}~\cite{Wask21}, and
\texttt{Quail}~\cite{HeusEtal17} (lines 26--28). Because the first two
are requirements of \texttt{UMAP}'s optional ``plot'' feature, they
will have already been installed (if necessary) by line 23, though possibly as
different versions than those specified in the onion comments on lines
26 and 28. If the installed and specified versions are the same, these
\texttt{smuggle} statements will function like standard \texttt{import}
statements to load the packages into the notebook's namespace. If they
differ, Davos will download the requested versions in place
of the installed versions, ensuring that they are used both in the notebook itself and by \texttt{UMAP} internally.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example6}
\end{center}
The onion comment in line 28 specifies that \texttt{Quail} should be
installed from a fork of its GitHub repository (\texttt{myfork}), in its state as of a specific commit (\texttt{6c847a4}).
This ability to load packages directly from remote (or local) Git repositories can
enable developers to more easily use forked or customized versions of other
packages in their code, even if those versions have not been
officially released. Targeting specific VCS references (e.g., commits, tags, etc.) can also provide even finer-grained control over smuggled package versions than is possible with traditional version specifiers.

@@

Lines 30--37 account for both of these possibilities programmatically:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
First, in line 30, we set the \texttt{davos.project} attribute to \texttt{None} to temporarily
%disable Davos's project isolation system and
allow installing smuggled packages outside of the notebook's project directory.
As noted in Section \ref{subsec:projects}, this is typically discouraged, as doing so can risk interfering with the user's Python environment if existing package versions are overwritten.
In this particular case, however, a combination of factors make this relatively safe and inconsequential.
First, the package we need to install directly into the notebook server environment (\texttt{widgetsnbextension}) is smuggled without an accompanying onion comment (line 34), meaning that Davos will not replace any version the user may already have installed.
Second, the package has no dependencies of its own, so if Davos does install it, no other packages could potentially be installed or updated as a side effect.
Third, the package itself provides no functionality outside of rendering Jupyter widgets, so its presence would not alter any other code's expected behavior.

>>>>

We designed Davos to occupy a ``sweet spot'' within this space.
Davos is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
Davos-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run a standard Jupyter
notebook. And like virtual environments, Davos provides a convenient
mechanism for fully specifying (and installing, as needed) a complete set of
Python dependencies, including specific package versions, which are contained
and isolated from the rest of the user's system.

@@

\item \texttt{.all\_projects}: This attribute contains a list of all Davos projects that exist on the user's local system (see Sec.~\ref{subsec:projects} for more information about Davos projects).
  Each item in this list is either a \texttt{Project} or \texttt{AbstractProject} instance.
  \texttt{Abstract\-Project}s represent notebook-specific projects whose associated notebooks no longer exist.
  They support the same functionality as \texttt{Project} objects (including methods for inspecting, renaming, and deleting them) and serve primarily to help users identify and clean up extraneous projects left behind after deleting Davos-enhanced notebooks (e.g., see Sec.~\ref{subsec:toplevel}).

@@

After reverting these two options (line 24) to their default values, we next \texttt{smuggle}
specific versions of three plotting packages:
\texttt{Matplotlib}~\cite{Hunt07}, \texttt{seaborn}~\cite{Wask21}, and
\texttt{Quail}~\cite{HeusEtal17} (lines 26--28). Because the first two
are requirements of \texttt{UMAP}'s optional ``plot'' feature, they
will have already been installed (if necessary) by line 23, though possibly as
different versions than those specified in the onion comments on lines
26 and 28. If the installed and specified versions are the same, these
\texttt{smuggle} statements will function like standard \texttt{import}
statements to load the packages into the notebook's namespace. If they
differ, Davos will download the requested versions in place
of the installed versions, ensuring that they are used both in the notebook itself and by \texttt{UMAP} internally.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example6}
\end{center}
The onion comment in line 28 specifies that \texttt{Quail} should be
installed from a fork of its GitHub repository (\texttt{myfork}), in its state as of a specific commit (\texttt{6c847a4}).
This ability to load packages directly from remote (or local) Git repositories can
enable developers to more easily use forked or customized versions of other
packages in their code, even if those versions have not been
officially released. Targeting specific VCS references (e.g., commits, tags, etc.) can also provide even finer-grained control over smuggled package versions than is possible with traditional version specifiers.

@@

Lines 30--37 account for both of these possibilities programmatically:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
First, in line 30, we set the \texttt{davos.project} attribute to \texttt{None} to temporarily
%disable Davos's project isolation system and
allow installing smuggled packages outside of the notebook's project directory.
As noted in Section \ref{subsec:projects}, this is typically discouraged, as doing so can risk interfering with the user's Python environment if existing package versions are overwritten.
In this particular case, however, a combination of factors make this relatively safe and inconsequential.
First, the package we need to install directly into the notebook server environment (\texttt{widgetsnbextension}) is smuggled without an accompanying onion comment (line 34), meaning that Davos will not replace any version the user may already have installed.
Second, the package has no dependencies of its own, so if Davos does install it, no other packages will be installed or updated as a side effect.
Third, the package itself provides no functionality outside of rendering Jupyter widgets, so its presence would not alter any other code's expected behavior.
